---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Conor Lacey"
date: "02-19-23"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
suppressWarnings(library(tidyverse))
library(dsbox)
library(mosaicData) 
```

### Exercise 1

```{r load-data}
staff <- read_csv("data/instructional-staff.csv")
```

I believe that there will be 11 rows of data when we convert this data to long format.

```{r long-conversion}
staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long
```


```{r employee-plot}
staff_long %>%
  ggplot(aes(x = year,
             y = value,
             group = faculty_type,
             color = faculty_type)) +
  geom_line()+
  labs(title = "Trends in Instructional Staff Employees",
       x = "Year",
       y = "Percentage of Hires",
       color = "Faculty Type")
```

To make it so that the graph highlights that the proportion of part-time faculty has gone up over the years in comparison to other faculty types, I would keep the part-time faculty line as its own line and consolidate the other faculty types into one line. The consolidated line should be lower than the part-time faculty line.

### Exercise 2

Remove this text, and add your answer for Exercise 1 here.
Add code chunks as needed.
Don't forget to label your code chunk.
Do not use spaces in code chunk labels.

### Exercise 3

...

Add exercise headings as needed.
